Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-22 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

\abstract{Single-cell RNA Velocity has dramatically advanced our ability
to model cellular differentiation and cell fate decisions. However,
current preprocessing choices and model assumptions often lead to errors
in assigning developmental trajectories. Here, we develop,
Pyro -Velocity, a Bayesian, generative, and multivariate RNA Velocity
model to estimate the uncertainty of cell future states. This approach
models raw sequencing counts with the synchronized cell time across all
expressed genes to provide quantifiable and improved information on cell
fate choices and developmental trajectory dynamics.}

@@

RNA Velocity is a powerful computational framework to estimate the time
derivative of gene expression
\citep{La_Manno2018-lj, Svensson2018-vk, Qiu2022-dj, Bergen2020-pj}. The
framework has been used to study developmental cell lineage trajectories
over a broad spectrum of biological processes, such as haematopoiesis
\citep{Qiu2022-dj}, pancreas development \citep{Bastidas-Ponce2019-lf},
and neural progenitors differentiation \citep{Bergen2020-pj}. Yet,
recent state-of-the-art RNA Velocity approaches such as velocyto and
scVelo \citep{La_Manno2018-lj} have several limitations. First, they may
predict developmental trajectories that do not exist in nature, as it
has been shown in the analysis of fully mature blood cells
\citep{Zheng2017-bz, Bergen2021-qz}. Second, data preprocessing steps
(e.g., selection of top variable genes, dimensionality reduction) affect
velocity estimates that are difficult to characterize and might lead to
artifactual predictions. In addition, non-linear spliced and unspliced
imputation procedures could distort phase portrait geometry and
recovered trajectories (Fig.~\ref{fig-model}a). Third, most RNA velocity
algorithms perform estimates independently for each gene, and therefore
cannot exploit the natural interdependence of gene expression. Fourth, a
unified description of the temporal relation between cells can only be
assigned based on postprocessing steps that aggregate heuristically the
gene-wise fits into a shared latent time (Fig.~\ref{fig-model}a).
Finally, current RNA velocity frameworks only output point estimates of
the velocity vector field and do not provide uncertainty estimation.
Importantly, the absence of statistical confidence estimates increases
false discoveries as it leaves users without a way to evaluate the
robustness of the methods' outputs. To overcome these limitations, we
developed a probabilistic generative model, named \emph{Pyro -Velocity},
that estimates RNA dynamics and infers cell trajectories using Pyro
\citep{Bingham2018-id}, a robust and efficient probabilistic programming
language.

@@

\includegraphics{figures/fig.pdf}

@@

\includegraphics{figures/fig.pdf}

@@

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right), \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)-\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align}

@@

\includegraphics{figures/fig.pdf}

@@

\includegraphics{figures/fig.pdf}

@@

\includegraphics{figures/fig.pdf}

@@

\includegraphics{figures/fig.pdf}

>>>>

\abstract{Single-cell RNA velocity has dramatically advanced our ability
to model cellular differentiation and cell fate decisions. However,
current preprocessing choices and model assumptions often lead to errors
in assigning developmental trajectories. Here, we develop,
Pyro -Velocity, a Bayesian, generative, and multivariate RNA Velocity
model to estimate the uncertainty of cell future states. This approach
models raw sequencing counts with the synchronized cell time across all
expressed genes to provide quantifiable and improved information on cell
fate choices and developmental trajectory dynamics.}

@@

RNA velocity is a powerful computational framework to estimate the time
derivative of gene expression
\citep{La_Manno2018-lj, Svensson2018-vk, Qiu2022-dj, Bergen2020-pj}. The
framework has been used to study developmental cell lineage trajectories
over a broad spectrum of biological processes, such as haematopoiesis
\citep{Qiu2022-dj}, pancreas development \citep{Bastidas-Ponce2019-lf},
and neural progenitors differentiation \citep{Bergen2020-pj}. Yet,
recent state-of-the-art RNA Velocity approaches such as velocyto and
scVelo \citep{La_Manno2018-lj} have several limitations. First, they may
predict developmental trajectories that do not exist in nature, as it
has been shown in the analysis of fully mature blood cells
\citep{Zheng2017-bz, Bergen2021-qz}. Second, data preprocessing steps
(e.g., selection of top variable genes, dimensionality reduction) affect
velocity estimates that are difficult to characterize and might lead to
artifactual predictions. In addition, non-linear spliced and unspliced
imputation procedures could distort phase portrait geometry and
recovered trajectories (Fig.~\ref{fig-model}a). Third, most RNA velocity
algorithms perform estimates independently for each gene, and therefore
cannot exploit the natural interdependence of gene expression. Fourth, a
unified description of the temporal relation between cells can only be
assigned based on postprocessing steps that aggregate heuristically the
gene-wise fits into a shared latent time (Fig.~\ref{fig-model}a).
Finally, current RNA velocity frameworks only output point estimates of
the velocity vector field and do not provide uncertainty estimation.
Importantly, the absence of statistical confidence estimates increases
false discoveries as it leaves users without a way to evaluate the
robustness of the methods' outputs. To overcome these limitations, we
developed a probabilistic generative model, named \emph{Pyro -Velocity},
that estimates RNA dynamics and infers cell trajectories using Pyro
\citep{Bingham2018-id}, a robust and efficient probabilistic programming
language.

@@

\includegraphics{figures/v2/Figure2.png}

@@

\includegraphics{figures/v2/Figure3.png}

@@

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
   \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align} In the equation, the subscript \(c\) is the cell dimension,
\(g\) is the gene dimension,
\(\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)\)
are the unspliced and spliced expression functions given the change of
time per cell and gene. \(\tau_{cg}\) represents the displacement of
time per cell and gene with \begin{align}
 \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
 & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
\end{align} in which \(t_c\) is the shared time per cell,
\({t_{0}^{(kcg)}}_g\) is the gene-specific switching time. Each cell and
gene combination has its transcriptional state
\(k_{cg} \in \{ 0, 1 \}\), where \(0\) indicates the activation state
and \(1\) indicates the expression state. Each gene has two switching
times for representing activation and repression: \({t_{0}^{(0)}}_g\) is
the first switching time corresponding to when the gene expression
starts to be activated, \({t_0^{(1)}}_g\) is the second switching time
corresponding to when the gene expression starts to be repressed. We
note that \(\alpha^{(1)}\) is shared for all the genes, while
\({\alpha^{(0)}}_g\) is learned independently for each gene.

The analytic solution of the differential equations to predict spliced
and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

To simplify these equations, consider the case when \(k_{cg} = 0\) and
\(\beta_g \neq \gamma_g\). Then, \begin{align}
u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
  \label{eq-sol-usimp} \\
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
  \nonumber\\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
\end{align} When \(k_{cg} = 0\) and \(\beta_g = \gamma_g\), then
\(u\left(\tau^{(0)}\right)\) has the same solution, and
\(s\left(\tau^{(0)}\right)\) becomes \begin{align}
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
\end{align} When \(k_{cg} = 1\) and \(\beta_g \neq \gamma_g\), then
\begin{align}
u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
\end{align} When \(k_{cg} = 1\) and \(\beta_g = \gamma_g\), then
\(u\left(\tau^{(1)}\right)\) has the same solution, and
\(s\left(\tau^{(1)}\right)\) becomes \begin{align}
s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
\end{align}

\subsection{Variational inference}\label{sec-methods-inference}

Given observations
\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align} but exact Bayesian inference is intractable in this model.
We use Pyro to automatically integrate out the local discrete latent
variables \(k\), which is defined as the cell and gene transcriptional
state (see above), and approximate the posterior over the remaining
latent variables using variational inference
\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
integrals into optimization problems that can be solved with
off-the-shelf tools. In variational inference, we maximize a tractable
bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
divergence
\[\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} p(\theta, \psi \vert \tilde{X}_{cg}) \right)\]
between a parametric family of tractable probability distributions q and
the intractable true posterior \begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align} We approximate our model's posterior distribution
\(p( \theta, \psi \vert
\tilde{X})\) with a tractable family of probability distributions
\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
learnable location parameter and a learnable low-rank covariance matrix.
We solve the resulting stochastic optimization problem using a version
of Automatic Differentiation Variational Inference (ADVI)
\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
differentiating a Monte Carlo estimate of the ELBO and update the
variational parameters using stochastic gradient ascent. Our
implementation and experiments use generic variational families and
Monte Carlo ELBO and gradient estimators provided by Pyro
\citep{Bingham2018-id}, an open source library for probabilistic machine
learning, and a version built into Pyro of the adaptive stochastic
gradient ascent algorithm Adam augmented with gradient clipping to
enhance numerical stability during training.

\subsection{Model training}\label{sec-methods-training}

For the pancreas, PBMC, and uni-fate and bi-fate LARRY single-cell data,
since the data dimension is relatively small, we run Model 1 and Model 2
with a minimum of 100 epochs and a maximum of 4000 epochs. For each
epoch, we input spliced and unspliced read counts from all the cells and
determine the convergence condition with an early stopping strategy in
which the patience is set to 45 and consumes one patience if the minimal
ELBO improvement of training data per epoch is lower than \(10^{-4}\) of
previous loss. The learning rate is set to be \(10^{-2}\) with a decay
rate of \(0.11/4000\) per epoch. For the multi-fate LARRY dataset, since
this is a large dataset with over \(4 \times
10^4\) cells, we use mini batches of cells to train Model 1 and Model 2
with a minimum of 100 epochs and a maximum of 1000 epochs. Specifically,
the batch size was set to 4000 cells for both models with an early
stopping patience 45 and consume one patience if the minimal ELBO
improvement of training data per epoch lower than \(10^{-3}\) of
previous loss. The learning rate is set to be \(10^{-2}\) with a decay
rate of \(0.11/1000\) per epoch. All models were trained on a machine
with an NVIDIA A100 GPU and the CentOS 7 operating system.

@@

\includegraphics{figures/v2/Supplementary-Figure2.png}

@@

\includegraphics{figures/v2/Supplementary-Figure3.png}

@@

\includegraphics{figures/v2/Supplementary-Figure4.png}

@@

\includegraphics{figures/v2/Supplementary-Figure5.png}
